---
title: "SA 2 - Comparing Training Day Errors on Drug Treatments and AD Status"
author: "Cuerdo, Naomi Hannah A."
date: "2024-12-08"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(car)
library(rstatix)
library(ggplot2)


```


```{r data, include=FALSE}
data <-read.csv("C:/Users/naomi/Downloads/Alzheimers-Mice-Data.csv")
```


```{r data con, include=FALSE}
data$AD_Status <-as.factor(data$AD_Status)
data$Treatment <- as.factor(data$Treatment)

```


# Introduction

This report examines the effects of Alzhimer's Disease (AD) status and drug treatment on training errors. Training errors were measured as the number of errors committed during a training session, recorded on a continuous scale. The independent variables included:

*- AD Status*: Categorized as "With AD" (coded as 1) and "Without AD" (coded as 2)

*- Treatment*: Drug treatment type categorized into four groups: "A" (coded as 1), "B" (coded as 2), "C" (coded as 3), and "D" (coded as 4)

Furthermore, this analysis explores the interaction between AD status and treatment type to understand their combined influence on training performance. 

# Method 

## Participants
The dataset consisted of a sample of experimental subjects assigned to two AD status groups and four treatment groups. The dependent variable was *training day errors*, measured as a continuous variable representing task performance.

## Procedure
Training errors were analyzed using a two-way (2B) ANOVA with two between-subject factors: AD status and treatment. This statistical method was chosen to assess the main effects of AD status and treatment, as well as their interaction effect, on training errors. 

Below are the assumptions:

Below are the assumptions:
*Assumption 1.* The dependent variable  is measured at the continuous level.
*Assumption 2.* The two independent variables consists of two or more categorical independent groups.
The AD Status is the dichotomous variable, whereas the Treatment is the polytomous variable.
*Assumption 3.* The independence of observations among each group of the independent variable or between the group themselves is observed.
*Assumption 4.* There should be no significant outliers in any cell of the design.
*Assumption 5.* The distribution of the dependent variable should be approximately normally distributed in every cell of the design.
*Assumption 6.* The variance of the dependent variable should be equal in every cell of the design.

## Hypothesis

**1.Main Effect on AD Status**

Null Hypothesis (H₀): There is no difference in training scores between individuals with AD and those without AD.

Alternative Hypothesis (H₁): There is a significant difference in training scores between individuals with AD and those without AD.

**2.Main Effect on Treatment**

Null Hypothesis (H₀): There is no difference in training across different treatment groups.

Alternative Hypothesis (H₁):There is a significant difference in training between at least two treatment groups.

**3.Interaction Effect between AD Status and Treatment**

Null Hypothesis (H₀):There is no interaction between AD status and treatment type in their effect on training scores. The effect of treatment is the same for individuals with and without AD.

Alternative Hypothesis (H₁):There is a significant main interaction effect on AD Status and treatment type on training scores.

# Descriptive Analysis

```{r head}
head(data)
```

```{r summary}
summary(data)

```
```{r}
standard_dev <- sd(data$Training, na.rm = TRUE)
print(sprintf("St dev: %.2f", standard_dev))
```
The dataset included 20 participants identified as "With AD" (coded as 1) and 20 as "Without AD" (coded as 2). There were 10 participants in each of the 4 treatment groups (1, 2, 3, 4)

The dependent variable, *Training Errors*, was measured on a continuous scale, with values ranging from 12 to 18 errors. The *mean* number of errors was ***M=14.72***, with a ***standard deviation (SD) = 1.75.***

***Thus, Assumptions 1 and 2 was achieved.***

```{r}
aggregate(Training ~ AD_Status + Treatment, data = data, FUN = base::mean)
```

The table above indicates that *Training Scores* vary across combinations of AD_Status and Treatment. Individuals with *AD (AD_Status = 1)* and *Treatment 3* reported the highest mean training scores *(M = 16.50)*, while individuals without AD *(AD_Status = 2)* in Treatment 1 reported the lowest mean scores *(M = 12.00)*.

Differences are also observed between *AD_Status* within each treatment category, such as higher training scores for those without AD *(M = 14.20)* compared to those with *AD (M = 12.00)* in *Treatment 1.*

Therefore, there is no overlap or duplication in data points within groups.

***Assumption 3 (independence of observations) was achieved.***

```{r}
ggplot(data, aes(x = AD_Status, y = Training, fill = Treatment)) +
  geom_boxplot() +
  labs(title = "Boxplot of Training Errors by AD Status and Treatment")
```

The figure shows that *With AD* participants have lower training scores, with a wider spread, compared to *Without AD participants*, who have higher scores. *Treatment 3* shows the highest training scores for both groups, with the *Without AD* group having a higher median. There are no significant outliers.


***Assumption 4 (no outliers) was achieved.***

```{r}
model <- aov(Training ~ AD_Status + Treatment, data = data,)
shapiro.test(residuals(model))
```

```{r}
shapiro_results <- data %>%
  group_by(AD_Status, Treatment) %>%
  summarise(
    shapiro_stat = shapiro.test(Training)$statistic,
    p_value = shapiro.test(Training)$p.value
  )

print(shapiro_results)
```

The normal distribution of the residual suggests that the model has appropriately captured the systemic patterns in training scores across AD status and treatment, therefore we fail to reject the null hypothesis (***W*** = 0.96359, ***p*** \> .05).

***Assumption 5 was achieved***

```{r}
levene_test(Training ~ AD_Status * Treatment, data = data)
```

From the table above, the training scores across the
independent variables show similar levels of variability (***F*** = 2.21, ***p*** \> .05). This means that the variances (or spread) of the training scores are *equal across the groups*. Because the p-value is greater than the significance level of 0.05, we *fail* to reject the null hypothesis that the variances are equal.

***Assumption 6 was achieved***

# Inferential Analysis

## Two-Way (2B) ANOVA

Now that all assumptions were checked, we can now proceed with the ANOVA Analysis.

```{r}
anova_training <- aov(Training ~ AD_Status * Treatment, data = data)
summary(anova_training)
```
From the given table above, it shows that there is a significant main effect of Treatment on the training scores (p=0.0197). Thus we proceed with a post hoc test:

```{r}
post_hoc <-TukeyHSD(anova_training, "Treatment", conf.level = 0.95)
print(post_hoc)
```
From the given table above, the only significant difference is between *Treatment 4 and Treatment 2*, wherein Treatment 4 has significantly lower training scores than Treatment 2. Other pairwise comparisons show no significant differences, suggesting that treatments 1, 2, and 3 do not differ significantly from each other in terms of training scores.

# APA Report

A two-way ANOVA was conducted to examine the effects of *AD Status* and *Treatment* on *Training Scores.* Assumptions of normality and homogeneity of variances were evaluated. The *Shapiro-Wilk test* for normality indicated that the residuals were approximately normally distributed *(W = 0.96359, p = 0.2219)*. Additionally, *Levene’s test* for homogeneity of variances was non-significant *(F(7, 32) = 2.21, p = 0.07)*, indicating equal variances across groups.

The results indicated a statistically significant main effect of Treatment on Training Scores *(F(3, 32) = 3.79, p = 0.02, η² = 0.26)*, suggesting that Training Scores significantly differ across the Treatment groups. Post hoc analysis revealed that Treatment 4 had significantly lower training scores compared to Treatment 2 *(p = 0.02)*. However, there were no significant differences between the other treatment pairs *(all p > 0.05)*.

No significant main effect of AD Status was found *(F(1, 32) = 1.22, p = 0.28, η² = 0.04)*, indicating that AD Status did not significantly influence Training Scores. Furthermore, there was no significant interaction effect between AD Status and Treatment *(F(3, 32) = 1.22, p = 0.32, η² = 0.10)*, suggesting that the effect of Treatment on Training Scores was not influenced by AD Status.

In summary, Treatment had a significant impact on Training Scores, but AD Status did not, and there was no significant interaction between the two variables.
